---
title: "Biomarkers of ASD"
subtitle: "If you want a subtitle put it here"
author: "List names here"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. In all, your report can be quite short. When it is complete, render and then push changes to your team repository.

```{r}
# load any other packages and read data here
library(tidyverse)
library(infer)
library(randomForest)
library(tidymodels)
library(modelr)
library(yardstick)
```

## Abstract: Nikhil

Write a brief one-paragraph abstract that describes the contents of your write-up.

## Dataset: Leena

Write a brief data description, including: how data were obtained; sample characteristics; variables measured; and data preprocessing. This can be largely based on the source paper and should not exceed 1-2 paragraphs.

## Summary of published analysis

Summarize the methodology of the paper in 1-3 paragraphs. You need not explain the methods in depth as we did in class; just indicate what methods were used and how they were combined. If possible, include a diagram that depicts the methodological design. (Quarto has support for [GraphViz and Mermaid flowcharts](https://quarto.org/docs/authoring/diagrams.html).) Provide key results: the proteins selected for the classifier and the estimated accuracy.

## Findings

Summarize your findings here. I've included some subheaders in a way that seems natural to me; you can structure this section however you like.

### Impact of preprocessing and outliers: Leena

Tasks 1-2

### Methodlogical variations: Nikhil/Sanchit/Shirley

a\) repeat the analysis on training partition

Multiple testing Method

```{r}
## MULTIPLE TESTING
####################

# function to compute tests
test_fn <- function(.df){
  t_test(.df, 
         formula = level ~ group,
         order = c('ASD', 'TD'),
         alternative = 'two-sided',
         var.equal = F)
}

ttests_out <- biomarker_clean %>%
  # drop ADOS score
  select(-ados) %>%
  # arrange in long format
  pivot_longer(-group, 
               names_to = 'protein', 
               values_to = 'level') %>%
  # nest by protein
  nest(data = c(level, group)) %>% 
  # compute t tests
  mutate(ttest = map(data, test_fn)) %>%
  unnest(ttest) %>%
  # sort by p-value
  arrange(p_value) %>%
  # multiple testing correction
  mutate(m = n(),
         hm = log(m) + 1/(2*m) - digamma(1),
         rank = row_number(),
         p.adj = m*hm*p_value/rank)

# select significant proteins
proteins_s1 <- ttests_out %>%
  slice_min(p.adj, n = 10) %>%
  pull(protein)

```

### Improved classifier: Nikhil/Sanchit/Shirley

```{r}

```

```{r}

```
